---
title: "Statistical Learning"
output: github_document
date: "2022-12-01"
---

```{r setup, include = FALSE}
library(tidyverse)
library(p8105.datasets)
library(glmnet)

set.seed(11)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Lasso

To illustrate the lasso, we’ll get data from a study of factors that affect birthweight. The code chunk below loads and cleans these data, converts to factors where appropriate, and takes a sample of size 200 from the result.

```{r}
bwt_df = 
  read_csv("./extra_topic_data/birthweight.csv") %>% 
  janitor::clean_names() %>%
  mutate(
    babysex = as.factor(babysex),
    babysex = fct_recode(babysex, "male" = "1", "female" = "2"),
    frace = as.factor(frace),
    frace = fct_recode(frace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4", "other" = "8"),
    malform = as.logical(malform),
    mrace = as.factor(mrace),
    mrace = fct_recode(mrace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4")) %>% 
  sample_n(200)
```

### Inputs for `glmnet`

To fit a lasso model, we’ll use `glmnet`. This package is widely used and broadly useful, but predates the tidyverse by a long time. The interface asks for an outcome vector `y` and a matrix of predictors `X`, which are created next. To create a predictor matrix that includes relevant dummy variables based on factors, we’re using `model.matrix` and excluding the intercept

```{r}
x = model.matrix(bwt ~ ., bwt_df)[,-1]
y = bwt_df$bwt
```

We fit the lasso model for each tuning parameter in a pre-defined grid `lambda`, and then compare the fits using cross validation. I chose this grid using the trusty “try things until it looks right” method; `glmnet` will pick something reasonable by default if you prefer that.

```{r}
lambda = 10^(seq(3, -2, -0.1))

lasso_fit =
  glmnet(x, y, lambda = lambda)

lasso_cv =
  cv.glmnet(x, y, lambda = lambda)

lambda_opt = lasso_cv$lambda.min
```

When lambda is big, we have nothing in our model, as lambda gets smaller, more covariates get added to our model. Deviance represents goodness-of-fit.

`cv.glmnet` does cross-validation for our models, and `lambda_opt` saves the optimal lambda value identified from cross-validation. The best `lambda` gives us the best prediction accuracy given the predictors we have for the outcome of interest.

### Plotting `lasso` models

The plot below shows coefficient estimates corresponding to a subset of the predictors in the dataset – these are predictors that have non-zero coefficients for at least one tuning parameter value in the pre-defined grid. As lambda increases, the coefficient values are shrunk to zero and the model becomes more sparse. The optimal tuning parameter, determined using cross validation, is shown by a vertical blue line.


```{r}
broom::tidy(lasso_fit) %>% 
  select(term, lambda, estimate) %>% 
  complete(term, lambda, fill = list(estimate = 0) ) %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate, group = term, color = term)) + 
  geom_path() + 
  geom_vline(xintercept = log(lambda_opt, 10), color = "blue", size = 1.2) +
  theme(legend.position = "none")
```

Penalty pulls coefficients towards 0 as lambda gets bigger. Eventually, all coefficients get shrunk to 0 as lambda gets big - the penalty outweighs the residual squares. The optimal lambda is indicated by the vertical line.

The next plot shows the CV curve itself. This is relatively shallow – having nothing at all in your model isn’t great, but you can get reasonable predictions from models that have “too many” predictors.

```{r}
broom::tidy(lasso_cv) %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate)) + 
  geom_point() 
```

The coefficients from the optimal model are shown below. Note we can apply `broom::tidy` to `lasso` outputs.

```{r}
lasso_fit = 
  glmnet(x, y, lambda = lambda_opt)

lasso_fit %>% broom::tidy()
```

To be clear, these don’t come with p-values and it’s really challenging to do inference. These are also different from a usual OLS fit for a multiple linear regression model that uses the same predictors: the lasso penalty affects these even if they’re retained by the model.

A final point is that on the full dataset, lasso doesn’t do you much good. With ~4000 datapoints, the relatively few coefficients are estimated well enough that penalization doesn’t make much of a difference in this example.

## Clustering: Pokémon

For ther first clustering example, we’ll use a dataset containing information about pokemon. The full dataset contains several variables (including some that aren’t numeric, which is a challenge for clustering we won’t address). To make results easy to visualize, we look only at hp and speed; a scatterplot is below.

```{r}
poke_df = 
  read_csv("./extra_topic_data/pokemon.csv") %>% 
  janitor::clean_names() %>% 
  select(hp, speed)
```

```{r}
poke_df %>% 
  ggplot(aes(x = hp, y = speed)) + 
  geom_point()
```

K-means clustering is established enough that it’s implemented in the base R `stats` package in the `kmeans` function. This also has a bit of an outdated interface, but there you go. The code chunk below fits the k-means algorithm with three clusters to the data shown above.

```{r}
kmeans_fit =
  kmeans(x = poke_df, centers = 3)
```

We can extract out cluster assignment using `broom::augment`. In particular, we’ll use `broom::augment` to add cluster assignments to the data, and plot the results.


```{r}
poke_df =
  broom::augment(kmeans_fit, poke_df)

poke_df %>% 
  ggplot(aes(x = hp, y = speed, color = .cluster)) +
  geom_point()
```

Clusters broadly interpretable, but this still doesn’t come with inference. Also, at boundaries between clusters, the distinctions can seem a bit … arbitrary. The code chunk below maps across a few choices for the number of clusters, and then plots the results.

```{r}
clusts =
  tibble(k = 2:4) %>%
  mutate(
    km_fit =    map(k, ~kmeans(poke_df, .x)),
    augmented = map(km_fit, ~broom::augment(.x, poke_df))
  )

clusts %>% 
  select(-km_fit) %>% 
  unnest(augmented) %>% 
  ggplot(aes(hp, speed, color = .cluster)) +
  geom_point(aes(color = .cluster)) +
  facet_grid(~k)
```


